Diagnoses and cell types

### 2025-10-08 analyses/infercnv-consensus-cell-type/exploratory-notebooks/reference-query-counts.Rmd
update notebook now that consensus are updated

```{r}
usable_df <- library_counts_df |>
  dplyr::group_by(diagnosis_group) |>
  dplyr::summarize(
    n_usable = sum(reference >= params$n_reference_threshold),
    total_group_libs = dplyr::n(), 
    frac_usable = round(n_usable / total_group_libs, 3) 
  ) |>
  dplyr::arrange(desc(total_group_libs)) |>
  dplyr::select(diagnosis_group, total_group_libs, n_usable, frac_usable)

usable_df
```
Except for a handful of "other solid tumors," our broad diagnosis groups are well-represented, which is nice to see.

@@

```{r}
usable_df |>
  dplyr::filter(frac_usable == 0)
```
It's not too surprising to see `Mixed phenotype acute leukemia` here since most immune cell types can't be included in the reference, and the other groups comprise very small numbers of libraries.

@@

```{r}
library_counts_df |>
  # consider usable only
  dplyr::filter(reference >= params$n_reference_threshold) |>
  dplyr::filter(query < params$n_reference_threshold)
```

There is one library with about 7x as many reference as query cells; let's look more closely:

```{r}
consensus_df |>
  dplyr::filter(scpca_library_id == "SCPCL000021") |>
  dplyr::count(consensus_annotation) |>
  dplyr::arrange(n)
```

This library indeed is dominated by immune cell types with very low tumor purity, but it still seems to have enough query cells to not be a problem and should be ok to run.

## Plots

@@

```{r}
usable_libraries <- library_counts_df |>
  dplyr::filter(reference >= params$n_reference_threshold) |>
  dplyr::pull(scpca_library_id)

plot_df <- query_reference_df |>
  dplyr::filter(scpca_library_id %in% usable_libraries) |>
  dplyr::left_join(validation_group_df, by = c("consensus_ontology", "consensus_annotation")) |>
  dplyr::select(
    scpca_project_id, 
    scpca_library_id, 
    barcodes, 
    cell_class, 
    validation_group_annotation
  ) |>
  dplyr::mutate(
    label = ifelse(
      cell_class == "query", 
      "query",
      validation_group_annotation
    ),
    # move "query" first
    label = forcats::fct_relevel(label, "query", after = 0)
  )
```

@@

Currently, _before_ incorporating `SCimilarity` into our consensus cell types, we are able to run about 44% of libraries through `inferCNV`. 
Most broad diagnosis groups are represented, except `Mixed phenotype acute leukemia` and a handful of other solid tumors.
All eligible projects except `SCPCP000011` (Retinoblastoma samples) are represented as well.

Note that we are including multiplexed libraries from project `SCPCP000009` here as well; 11 libraries from that project from are eligible, most of which are multiplexed.
These libraries would effectively be using pooled references in `inferCNV`, since we did not demultiplex libraries.

>>>>

```{r}
usable_counts_df <- library_counts_df |>
  dplyr::group_by(diagnosis_group) |>
  dplyr::summarize(
    n_usable = sum(reference >= params$n_reference_threshold),
    total_group_libs = dplyr::n(), 
    frac_usable = round(n_usable / total_group_libs, 3) 
  ) |>
  dplyr::arrange(desc(total_group_libs)) |>
  dplyr::select(diagnosis_group, total_group_libs, n_usable, frac_usable)

usable_counts_df
```
Except for a handful of "other solid tumors," our broad diagnosis groups are well-represented, which is nice to see.

@@

```{r}
usable_counts_df |>
  dplyr::filter(frac_usable == 0)
```
It's not too surprising to see `Mixed phenotype acute leukemia` here since most immune cell types can't be included in the reference, and the other groups comprise very small numbers of libraries.

@@

```{r}
usable_libraries_df <- library_counts_df |>
  dplyr::filter(reference >= params$n_reference_threshold)

fewer_query_df <- usable_libraries_df |>
  dplyr::filter(query < params$n_reference_threshold)

fewer_query_df
```

There are five library with an order of magnitude more  7x as many reference as query cells; let's look more closely at the validation groups present to get a sense.

```{r}
consensus_df |>
  dplyr::filter(scpca_library_id %in% fewer_query_df$scpca_library_id) |>
  dplyr::left_join(validation_group_df, by = c("consensus_ontology", "consensus_annotation")) |>
  dplyr::count(scpca_library_id, validation_group_annotation) |>
  dplyr::arrange(scpca_library_id)
```

These librares are dominated by immune cell types with very few Unknown/tumor cells, but it's still worth running them through `inferCNV` while being aware that we _may_ experience issues with these samples.

We should also check the eligible `SCPCP000009` libraries to see if they are multiplexed:

```{r}
library_diagnosis_df |>
  dplyr::filter(scpca_project_id == "SCPCP000009", scpca_library_id %in% usable_libraries_df$scpca_library_id) |>
  dplyr::count(is_multiplexed)
```

## Plot cell types

@@

```{r}
plot_df <- query_reference_df |>
  dplyr::filter(scpca_library_id %in% usable_libraries_df$scpca_library_id) |>
  dplyr::left_join(validation_group_df, by = c("consensus_ontology", "consensus_annotation")) |>
  dplyr::select(
    scpca_project_id, 
    scpca_library_id, 
    barcodes, 
    cell_class, 
    validation_group_annotation
  ) |>
  dplyr::mutate(
    label = ifelse(
      cell_class == "query", 
      "query",
      validation_group_annotation
    ),
    # move "query" first
    label = forcats::fct_relevel(label, "query", after = 0)
  )
```

@@

We are able to run about 51% of libraries through `inferCNV`, based on consensus cell types derived from `SingleR`, `CellAssign`, and `SCimilarity` labels.
Most broad diagnosis groups are represented, except `Mixed phenotype acute leukemia` (n=18) and a handful of other solid tumors (n=3 among Gastrointestinal stromal tumor and Hepatoblastoma).
All eligible projects are represented as well.

Note that we are including multiplexed libraries from project `SCPCP000009` here as well; 12 libraries from that project from are eligible, and all but of of those are multiplexed.
These libraries would effectively be using pooled references in `inferCNV`, since we did not demultiplex libraries.
